---
title: "Forecasting monthly wind generation in Spain"
author: "Víctor Rodríguez Albendea"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(forecast)
library(lmtest)
source("../src/forecasting_helpers.R")
```

# Project note

This is the cleaned public version of an academic forecasting exercise.
The original local coursework contained multiple drafts, course files and workbook-based inputs.
For portfolio use, the workflow has been simplified to a reproducible report driven by a local CSV.

# Load the series

```{r}
series_path <- "../examples/wind_generation_spain_monthly_2020_2024.csv"
ts_wind <- load_wind_series(series_path)
split_data <- train_test_split_ts(ts_wind, holdout = 5)

train_ts <- split_data$train
test_ts  <- split_data$test

ts_wind
```

# Exploratory view

```{r}
plot(
  ts_wind,
  main = "Monthly wind generation in Spain (2020-2024)",
  xlab = "Year",
  ylab = "Wind generation (MW)",
  col = "#0f5a8c",
  lwd = 2
)
```

```{r}
decomp <- stl(ts_wind, s.window = "periodic")
plot(decomp, main = "STL decomposition")
```

# Smoothing baselines

## Simple exponential smoothing

```{r}
fit_ses <- HoltWinters(train_ts, beta = FALSE, gamma = FALSE)
pred_ses <- predict(fit_ses, n.ahead = length(test_ts))
rmse_ses <- holdout_rmse(test_ts, pred_ses)
rmse_ses
```

## Holt trend model

```{r}
fit_holt <- HoltWinters(train_ts, gamma = FALSE)
pred_holt <- predict(fit_holt, n.ahead = length(test_ts))
rmse_holt <- holdout_rmse(test_ts, pred_holt)
rmse_holt
```

## Holt-Winters

```{r}
fit_hw <- hw(train_ts, seasonal = "additive")
pred_hw <- forecast(fit_hw, h = length(test_ts))
rmse_hw <- holdout_rmse(test_ts, pred_hw$mean)
rmse_hw
```

# ARIMA modelling

The original academic work explored Box-Cox transformation, seasonal differencing and manual ARIMA identification.
This public version keeps the same idea while using a compact modelling path.

```{r}
lambda_bc <- BoxCox.lambda(train_ts)
lambda_bc
```

```{r}
fit_arima <- auto.arima(
  train_ts,
  seasonal = TRUE,
  lambda = lambda_bc,
  stepwise = FALSE,
  approximation = FALSE
)

fit_arima
```

```{r}
pred_arima <- forecast(fit_arima, h = length(test_ts))
rmse_arima <- holdout_rmse(test_ts, pred_arima$mean)
rmse_arima
```

# Holdout comparison

```{r}
results <- data.frame(
  model = c("SES", "Holt", "Holt-Winters", "ARIMA"),
  holdout_rmse = c(rmse_ses, rmse_holt, rmse_hw, rmse_arima)
)

results[order(results$holdout_rmse), ]
```

# Residual diagnostics

```{r}
checkresiduals(fit_arima)
```

# Forecast preview

```{r}
future_forecast <- forecast(fit_arima, h = 12)
plot(
  future_forecast,
  main = "12-month forecast preview from the selected ARIMA model",
  xlab = "Year",
  ylab = "Wind generation (MW)"
)
```

# Remarks

- The public version focuses on clarity and reproducibility.
- It does not include the original course workbook, Word templates or lecture notes.
- The staged CSV is enough to preserve the modelling narrative of the project.

